DRAFT -- Discriminating Seagrasses From Green Macroalgae in European Intertidal areas using high resolution multispectral drone imagery -- DRAFT

### 2024-03-22 index.qmd
pierre's comments

  Coastal areas support seagrass meadows, which offer crucial ecosystem services including erosion control and carbon sequestration. However, these areas are increasingly impacted by human activities, leading to seagrass decline and habitat fragmentation. In situ surveys, traditionally performed to monitor these ecosystems face limitations on temporal and spatial coverage, particularly in intertidal zones, prompting the use of satellite data within monitoring programs. Yet, satellite remote sensing struggles with spatial and spectral resolution, making it difficult to discriminate seagrass from other macrophytes in highly heterogenous meadows. Drone images at a very high spatial resolution offer a promising solution to address challenges related to spatial heterogeneity and intrapixel mixture. This study focuses on using drone acquisitions with a ten spectral band sensor mirroring those of Sentinel-2, for mapping intertidal macrophytes and effectively discriminating between seagrass and green macroalgae. Nine drone flights were conducted at two different altitudes (12m and 120m) across heterogeneous intertidal European habitats in France and Portugal. Low altitude flights were used to train a Deep Learning classifier based on Neural Networks to discrimintate among five taxonomic classes of intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae) and benthic Bacillariophyceae (Diatoms). Classification of drone imagery resulted in an overall accuracy of 94% across all the sites and images, covering a total area of 467 000 m². The model exhibited an accuracy of 96.4% in identifying seagrass. This work showed that although they share the same pigmentary composition, seagrass and green algae can be discriminated using a multispectral sensor with ten spectral bands at 444, 475, 531, 560, 650, 668, 705, 717, 740 and 840 nm. The algorithm should now be adapted for Sentinel-2 to see if this discrimination is still operational at a coarser spatial resolution. 

@@

Coastal areas are vital hotspots for marine biodiversity, with intertidal seagrass meadows playing a crucial role at the interface between the land and oceans [@unsworth2022]. Seagrass meadows provide a myriad of ecosystem services to humanity, including carbon sequestration, oxygen production, protection against sea-level rise and coastline erosion, and limitation of eutrophication. They serve as vital habitats for a diverse array of marine and terrestrial species, providing living, breeding, and feeding grounds [@gardner2018 ; @Zoffoli2022 ; @jankowska2019]. Due to the concentration of human activities in coastal zones, seagrass meadows are directly exposed to and impacted by anthropogenic pressures. Global regression and fragmentation are currently observed due to climate change, diseases, urbanization, land reclamation, dredging, competition with alien species, and reduction in water quality [@nguyen2021 ; @soissons2018 ; @orth2006 ; @lin2018 ; @duffy2019 ; @rasheed2011long ; @chefaoui2018dramatic]. Both habitat fragmentation and reduction, in turn, can severely copromise the effectiveness of ecosystem services provided by seagrass meadows. While improvements in water quality have been recently reported in Europe, allowing an overall recovery of seagrass ecosystems at local scale, many coastal waters worldwide are still subjected to strong eutrophication processes [@deSantos2019 ; @Zoffoli2021]. Coastal eutrophication has been ass ociated to excessive accumulation of green macroalgae, so-called green tides (@devlin2023nutrients). Green tides produce shade and suffocation over seagrass individuals, thus threatening the health of seagrass ecosystems [@wang2022].

@@

While satellite remote sensing (RS) provides temporally consistent observations over large spatial scales, its utilization over intertidal areas is limited by several constraints. Satellite missions with a high temporal resolution (e.g. daily MODIS observation) are limited by a coarse spatial resolution (\>100m) over these restricted extension areas. Missions with a high spatial resolution such as Sentinel-2 (10m) or Landsat8/9 (30m) can be limited by low spectral resolution. The limited number of spectral bands is indeed a challenge to accurately discriminate seagrass from others co-existing macrophytes. In particular Chlorophyceae (green algae) and marine Magnoliopsida (seagrass) share the same pigment composition [@ralph2002 ; @Douay2022]. Therefore, to someone not specialized in the field, their spectral signatures may appear to be alike [@Davies2023 ; @bannari2022]. Recently, using advanced machine-learning algorithms trained with a large hyperspectral library of more than 300 field reflectance spectra, @Davies2023 demonstrated that it was possible to discriminate Magnoliopsida from Chlorophyceae using radiometric data acquired at Sentinel-2 's spectral resolution. However the application of this approach to satellite RS remains to be validated. Moreover patches of green algae can develop at small spatial scales that are not observable using non-commercial satellite imagery [@tuya2013], especially during the initial stage of a green tide.

Drones can potentially fill the gaps left by satellite RS and in situ measurements, due to their ability to provide spatially-explicit observations at very high spatial resolutions (pixel size from cm to mm) while capturing data at multi-spectral resolution [@fairley2022drone ; @oh2017use]. The versatility of drones allows for their application across a diverse thematic range , from coastal zone management [@adade2021 ; @casella2020 ; @angnuureng2022] to mapping species distribution [@joyce2023 ; @tallam2023 ; @Roca2022 ; @Roman2021 ; @Brunier2022Topographic]. However, when applied to coastal habitat mapping, most case studies are limited to a single flight, restricting the generalizability of their application over wider geographical scales [@Roman2021 ; @collin2019improving ; @rossiter2020uav ; @Brunier2022Topographic]. The present study aimed at analyzing the potential of multispectral drone RS to map intertidal macrophytes, with a particular focus on discriminating Magnoliopsida and Chlorophyceae (Seagrass and Green Algae respectively). Ten drone flights were performed over soft-bottom intertidal areas along two European countries (France and Portugal), covering a wide range of habitats, from monospecific seagrass meadows to meadows mixed with green, or red macroalgae. A deep learning algorithm was trained and validated for macrophyte discrimination, emphasizing applicability across diverse sites without a loss of prediction accuracy.

@@

```{r Figure1}
#| cache: false
#| echo: false
#| warning: false
#| fig-cap: >
#|   Location of drone flights in France and Portugal. A: Gulf of Morbihan (Two sites), B: Bourngeuf Bay (Two sites), C: Ria de Aveiro Coastal Lagoon (Three sites). Green represents intertidal areas. 
#| label: fig-map
#| out-width: "100%"

@@

At each location, a DJI Matrice 200 quadcopter drone equipped with a Micasense RedEdge Dual MX multispectral camera was flown to take 1.2 million pixel reflectance photographs with ten spectral bands ranging from the blue to the near infrared (NIR): 444, 475, 531, 560, 650, 668, 705, 717, 740 and 840 nm. To ensure consistent lighting conditions across flight paths, the drone's trajectory was aligned to maintain a solar azimuth angle of 90 degrees. An overlap of 70% and 80% (side and front respectively) between each image was set for each flight. A downwelling light sensor (DLS2) was used to acquire irradiance data concomitantly with the camera measurements. Raw data were calibrated in reflectance using a calibration panel reflective at \~50% provided by the manufacturer. Across all sites, flights were made at two different altitudes : 12 m or/and 120 m. (@tbl-flights).

@@

```{r Table1}
#| echo: false
#| warning: false
#| label: tbl-flights
#| tbl-cap: List of drone flight, summarising the date, the altitude and the purpose of each flight. 12 m  and 120 m flights have a spatial resolution of 8 and 80 mm respectively.
#| cache: true

@@

Before each flight, targets used as Ground Control Points (GCPs) were distributed over the study site and georeferenced with a Trimble © Geo XH 6000 differential GPS (dGPS). GCPs were used to correct georeferencing imprecision of orthomosaics with an horizontal and vertical accuracy of 10cm. dGPS was also used to georeference quadrats of 0.25 m² assessing the presence or absence of five key taxonomic classes of intertidal vegetation : Bacillariophyceae (Unicellular benthic diatoms forming biofilms at the sediment surface during low tide), Phaeophyceae (brown macroalgae), Magnoliopsida (dwarf seagrass), Chlorophyceae (green macroalgae) and Rhodophyceae (red macroalgae) (@fig-vegetation). Pictures of each quadrat were uploaded online to the Global Biodiversity Information Facility (GBIF) platform [@BedeGbif], a public open portal to store and share biodiversity data. Each photograph was also processed to estimate the percent cover of each type of vegetation using an image processing software (imageJ). For all quadrats, the hyperspectral reflectance signatures of each vegetation class was measured using an ASD FieldSpec HandHeld 2 spectroradiometer, which acquires reflectance between 325 and 1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures from quadrats serve for dual purposes: they validate the radiometric calibration of drone data and contribute to error reduction in photo interpretation. The workflow developed in this study is presented in @fig-workflow.

@@

```{r Table2}
#| echo: false
#| warning: false
#| tbl-cap: Vegetation Classes of the model and the number of pixels used to train and validate each class
#| label: tbl-validationPX
#| cache: true

@@

A dataset containing photo-interpreted drone reflectance pixels was built to train a Neural Network model with 2 hidden layers. The training pixels were categorized into seven different classes, representing the various habitats encountered at the different study sites: Sediment, Water, Chlorophyceae, Magnoliopsida, Bacillariophyceae, Phaeophyceae and Rhodophyceae. Only low-altitude flights (@tbl-flights) were used for training because their 8 mm spatial resolution allowied to avoid spectral sub-pixel mixing and to accurately identify vegetation classes. More than `r my_comma(round(sum(df_training$Training_Pixels),-3))` pixels at 8 mm resolution from the 3 training flights were used to train the model (@tbl-validationPX). Twenty one variables were used by the model as predictors: the ten raw spectral bands of the Micasense RedEdge Dual MX multispectral camera (ranging from 444 nm to 840 nm), the same ten spectral bands standardized using a min/max transformation ([@eq-std] ; @Cao2017) and the Normalized difference vegetation index (NDVI, @eq-ndvi). Standardisation of spectral bands is commonly used to eliminate the scaling differences between spectra and to limit the effect of biomass on the spectra shape [@Douay2022 ; @Davies2023].

@@

The model was applied to all flights at both 12 and 120 m of altitude. In situ information on georeferenced class type and percent cover collected during each flight was used to assess the model accuracy. These images were used to construct a validation dataset indicating the presence or absence of each class. Additionally to the quadrat-based validation dataset, polygons of each class were photo interpreted in order to increase the number of pixels of the validation dataset. A confusion matrix, along with precision metrics such as global accuracy, sensitivity, specificity, and Kappa coefficient, was generated for each sites. All validation matrices were then aggregated to create a unique matrix . Altogether, a total of `r my_comma(round(count_validation_pixel,-3))` pixels was used to validate the model, thus providing a geographically robust validation dataset.

@@

The high-altitude flight over Gafanha covered a total area of approwimately 1 km² (@fig-GafHigh). A channel contouring a small island was masked in the prediction map. Most of the intertidal area was classified as Magnoliopsida by the model, including seagrass patches with discolored leaves. Only a few pixels were classified as Chlorophyceae at this spatial scale. Furthermore, the area that was classified as Bacillariophyceae in the low-altitude flight remained mostly classified as such in the high-altitude flight, though some pixels were classified as Magnoliopsida. Patches of Rhodophyceae were correctly classified. In the northern part of the site and near the land eadges, patches of the schorre angiosperm *Spartina sp.* were misclassified, either as Magnoliopsida or as Phaeophyceae.

@@

Amoung the high altitude flights, the one acquired over the inner lagoon of the Ria de Aveiro covered the largest area with approximately 1.5 km² (@fig-Boat). This site was dominated by seagrass and the red macroalgae *Gracilaria vermicullophyla*. The classification provided consistent results, with a patchy seagrass meadow mixed with red macroalgae on the eastern part of the site. As shown in the zoom (@fig-Boat), the edges of the meadow were colonised by green macroalgae

@@

```{r VIP Plot}
#| cache: false
#| echo: false
#| warning: false
#| fig-cap: Variable Importance of the Neural Network Classifier for each taxonomic class. The bigger the slice, the more important the variable for prediction of each class. The right plot shows the drone raw and standardised reflectance spectra of each class. Each slice represents the VIP of both raw and standardised reflectance combined.      
#| label: fig-VIP
#| fig-width: 10
#| fig-height: 7
#| out-width: "100%"

@@

```{r Upscaling Plot}
#| cache: false
#| echo: false
#| warning: false
#| fig-cap: Kernel density plot showing the proportion of pixel well classified based on the percent cover of the class in high altitude flight pixels of Gafanha, Portugal. Each subplot shows all the pixels of the same classes on the high altitude flight. Percent cover of classes was retrieved using the result of the classification of the low altitude flight of Gafanha, Portugal. The vertical dashed line shows the 0.85 probability of the model. Everything on the right of this line has a probability higher then 0.85 and everything on the left of this line has a lower probability.
#| label: fig-upscaling
#| fig-width: 10
#| fig-height: 7
#| out-width: "90%"

@@

The primary objective of this study was to develop a method for the accurate classification of macrophytes on intertidal mudflats, specifically focusing on distinguishing between Chlorophyceae (green macroalgae) and marine Magnoliopsida (seagrasses) using multispectral drone data. The ability to differentiate between various types of vegetation, as demonstrated in this study, plays a critical role in ecological monitoring and coastal management. By distinguishing between seagrasses and macroalgae, our approach facilitates targeted conservation strategies, enabling more effective preservation and restoration efforts in coastal ecosystems. The discrimination of seagrasses from green macroalgae presents significant challenges [@oiry2021using ; @bannari2022 ; @veettil2020opportunities], primarily due to the similarities on their spectal signatures in the visible region of the eletromagnetic spectra due to shared pigment [@mckenzie2020global ; @fig-Pigm], but also by the frequent spatial mixing of these macrophytes. Besides chlorophyll-a that is common to all vegetation types, these two green macrophytes have in common other photosynthetic pigments such as chlorophyll-b, and accessory carotenoid pigments such as zeaxantin, lutein and neoxanthin (@fig-Pigm).

@@

This spectral overlap is further complicated by the intermingling of seagrasses and green macroalgae in the same spatial locations, posing a significant challenge for remote sensing aimed at accurately mapping and monitoring coastal ecosystems. Our study addresses these complexities by using high-resolution drone imagery with 10 spectral bands. @Davies2023 have shown that having at least eight spectral bands ranging between 500 nm to 850 nm including a green band at 530 nm and another one in the RedEdge region at 730 nm was crucial to accurately discriminate amoung green macrophytes.

@@

::: column
Meeting these two criteria, the Micasense RedEdge-MX DUAL camera used in this study, enabled the classifier to achieve `r round(validation_metrics_green$overall[[1]],2)*100`% of accuracy between these two classes (@fig-ValidationGreen). Even if the pigment composition of green macrophytes is similar, differences in the spectral shape can still be observed (@fig-vegetation). Several factors can explain these differences such has different concentration and/or pigment ratios [@bargain2013seasonal], different cellular organisation and the disposition of the plant over the sediment surface [@beach1997vivo ; @kirk1994light ; @hedley2018influence].
:::
:::

The VIP analysis of the Neural Network model (@fig-VIP) identified that the band at 531 nm was the most important one for accurately identifying Chlorophyceae. In fact, at this wavelength, Chlorophyceae exhibited the highest reflectance (raw and standardised) among all other classes, highlighting the potential differences in accessory pigment ratios (i.e. carotenoid) between seagrasses and green macroalgae [@repolho2017seagrass].

Concerning Phaeophyceae, the thick cell walls of these macroalgae [@charrier2021growth] make it more reflective in the infrared part of the spectra [@Slaton2001] whereas the presence of Fucoxanthin and Zeaxanthin result in a low reflectance in the visible region (@fig-Pigm ; @fig-VIP). These two key features have been identified by the Neural Network as the two principal predictors to accurately identify brown algae (@fig-VIP). Similarly, the presence of phycoerythrin and phycocyanin in Rhodophyceae contributes to the lowest reflectance among all classes in the spectral range of 560 to 615 nm (@fig-VIP). Indeed the band at 560 nm has been identified as important for identifying this class likely due to phycoerythrin absorption at this wavelength (**REF**). Regarding Bacillariophyceae, the VIP analysis (@fig-VIP) indicated that 475 nm was the most important predictor for this class. This is not a usual diagnostic wavelength for Bacillariophyceae which is mainly identified by the presence of fucoxanthin and chlorophyll c (@fig-Pigm), but it correspond to the absorption of diadinoxanthin an accessory pigment characteristic of this class [@meleder2003spectrometric]. Furthermore, it is the vegetation with the lowest concentration of chlorphyll-a, pigment absorbing light both in the blue and the red. The transparency of Bacillariophyceae makes the reflectance of the sediment part of the overall reflectance of Bacillariophyceae, further explaining the high reflectance in the blue.

@@

While comparing the reflectance of both altitudes (12 m and 120 m), it was observed @fig-CompareRef that there is an underestimation of the infrared part of the spectra in the high-altitude dataset. Such disparity in infrared reflectance may stem from temporal differences between the flights, possibly resulting in a slightly drier intertidal area and consequently higher infrared reflectance. This disparity poses an issue for the methodology followed in the present study, relying solely on one flight height for training. To address this issue, we employed min/max standardized reflectance spectra as predictors for the model (@eq-std ; [@Cao2017]). This approach allowed us to eliminate the slight reflectance difference between the flights (@fig-CompareRef B) and to focus on the shape of the spectra in the visible part of the electromagnetic spectra, where different pigmentation are associated to taxonomic dignostic features. This was a key feature in building a model that could reliably predict vegetation across geographical sites and seasons. It enabled consistent prediction of vegetation classes across variations in biomass and variability in light conditions [@fyfe2003spatial ; @COSTA2021107018 ; @piaser2023impact].

>>>>

  Coastal areas support seagrass meadows, which offer crucial ecosystem services including erosion control and carbon sequestration. However, these areas are increasingly impacted by human activities, leading to seagrass decline and habitat fragmentation. In situ surveys, traditionally performed to monitor these ecosystems face limitations on temporal and spatial coverage, particularly in intertidal zones, prompting the use of satellite data within monitoring programs. Yet, satellite remote sensing struggles with spatial and spectral resolution, making it difficult to discriminate seagrass from other macrophytes in highly heterogenous meadows. Drone images at a very high spatial resolution offer a promising solution to address challenges related to spatial heterogeneity and intrapixel mixture. This study focuses on using drone acquisitions with a ten spectral band sensor mirroring those of Sentinel-2, for mapping intertidal macrophytes and effectively discriminating between seagrass and green macroalgae. Nine drone flights were conducted at two different altitudes (12m and 120m) across heterogeneous intertidal European habitats in France and Portugal. Low altitude flights were used to train a Deep Learning classifier based on Neural Networks to discriminate among five taxonomic classes of intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae) and benthic Bacillariophyceae (Diatoms). Classification of drone imagery resulted in an overall accuracy of 94% across all the sites and images, covering a total area of 467 000 m². The model exhibited an accuracy of 96.4% in identifying seagrass. This work showed that although they share the same pigmentary composition, seagrass and green algae can be discriminated using a multispectral sensor with ten spectral bands at 444, 475, 531, 560, 650, 668, 705, 717, 740 and 840 nm. The algorithm should now be adapted for Sentinel-2 to see if this discrimination is still operational at a coarser spatial resolution. 

@@

Coastal areas are vital hotspots for marine biodiversity, with intertidal seagrass meadows playing a crucial role at the interface between the land and oceans [@unsworth2022]. Seagrass meadows provide a myriad of ecosystem services to humanity, including carbon sequestration, oxygen production, protection against sea-level rise and coastline erosion, and limitation of eutrophication. They serve as vital habitats for a diverse array of marine and terrestrial species, providing living, breeding, and feeding grounds [@gardner2018 ; @Zoffoli2022 ; @jankowska2019]. Due to the concentration of human activities in coastal zones, seagrass meadows are directly exposed to and impacted by anthropogenic pressures. Global regression and fragmentation are currently observed due to climate change, diseases, urbanization, land reclamation, dredging, competition with alien species, and reduction in water quality [@nguyen2021 ; @soissons2018 ; @orth2006 ; @lin2018 ; @duffy2019 ; @rasheed2011long ; @chefaoui2018dramatic]. Both habitat fragmentation and reduction, in turn, can severely compromise the effectiveness of ecosystem services provided by seagrass meadows. While improvements in water quality have been recently reported in Europe, allowing an overall recovery of seagrass ecosystems at local scale, many coastal waters worldwide are still subjected to strong eutrophication processes [@deSantos2019 ; @Zoffoli2021]. Coastal eutrophication has been associated to excessive accumulation of green macroalgae, so-called green tides (@devlin2023nutrients). Green tides produce shade and suffocation over seagrass individuals, thus threatening the health of seagrass ecosystems [@wang2022].

@@

While satellite remote sensing (RS) provides temporally consistent observations over large spatial scales, its utilization over intertidal areas is limited by several constraints. Satellite missions with a high temporal resolution (e.g. daily MODIS observation) are limited by a coarse spatial resolution (\>100m) over these restricted extension areas. Missions with a high spatial resolution such as Sentinel-2 (10m) or Landsat8/9 (30m) can be limited by low spectral resolution. The limited number of spectral bands is indeed a challenge to accurately discriminate seagrass from others co-existing macrophytes. In particular Chlorophyceae (green algae) and marine Magnoliopsida (seagrass) share the same pigment composition [@ralph2002 ; @Douay2022]. Therefore, to someone not specialized in the field, their spectral signatures may appear to be alike [@Davies2023 ; @bannari2022]. Recently, using advanced machine-learning algorithms trained with a large hyperspectral library of more than 300 field reflectance spectra, @Davies2023 demonstrated that it was possible to discriminate Magnoliopsida from Chlorophyceae using radiometric data acquired at Sentinel-2 's spectral resolution. However, the application of this approach to satellite RS remains to be validated. Moreover, patches of green algae can develop at small spatial scales that are not observable using non-commercial satellite imagery [@tuya2013], especially during the initial stage of a green tide.

Drones can potentially fill the gaps left by satellite RS and in situ measurements, due to their ability to provide spatially-explicit observations at very high spatial resolutions (pixel size from cm to mm) while capturing data at multi-spectral resolution [@fairley2022drone ; @oh2017use]. The versatility of drones allows for their application across a diverse thematic range, from coastal zone management [@adade2021 ; @casella2020 ; @angnuureng2022] to mapping species distribution [@joyce2023 ; @tallam2023 ; @Roca2022 ; @Roman2021 ; @Brunier2022Topographic]. However, when applied to coastal habitat mapping, most case studies are limited to a single flight, restricting the generalizability of their application over wider geographical scales [@Roman2021 ; @collin2019improving ; @rossiter2020uav ; @Brunier2022Topographic]. The present study aimed at analyzing the potential of multispectral drone RS to map intertidal macrophytes, with a particular focus on discriminating Magnoliopsida and Chlorophyceae (Seagrass and Green Algae respectively). Ten drone flights were performed over soft-bottom intertidal areas along two European countries (France and Portugal), covering a wide range of habitats, from monospecific seagrass meadows to meadows mixed with green, or red macroalgae. A deep learning algorithm was trained and validated for macrophyte discrimination, emphasizing applicability across diverse sites without a loss of prediction accuracy.

@@

```{r Figure1}
#| cache: false
#| echo: false
#| warning: false
#| fig-cap: >
#|   Location of drone flights in France and Portugal. A: Gulf of Morbihan (Two sites), B: Bourgneuf Bay (Two sites), C: Ria de Aveiro Coastal Lagoon (Three sites). Green represents intertidal areas. 
#| label: fig-map
#| out-width: "100%"

@@

At each location, a DJI Matrice 200 quadcopter drone equipped with a Micasense RedEdge Dual MX multispectral camera was flown to take 1.2-million-pixel reflectance photographs with ten spectral bands ranging from the blue to the near infrared (NIR): 444, 475, 531, 560, 650, 668, 705, 717, 740 and 840 nm. To ensure consistent lighting conditions across flight paths, the drone's trajectory was aligned to maintain a solar azimuth angle of 90 degrees. An overlap of 70% and 80% (side and front respectively) between each image was set for each flight. A downwelling light sensor (DLS2) was used to acquire irradiance data concomitantly with the camera measurements. Raw data were calibrated in reflectance using a calibration panel reflective at \~50% provided by the manufacturer. Across all sites, flights were made at two different altitudes : 12 m or/and 120 m. (@tbl-flights).

@@

```{r Table1}
#| echo: false
#| warning: false
#| label: tbl-flights
#| tbl-cap: List of drone flight, summarizing the date, the altitude and the purpose of each flight. 12 m  and 120 m flights have a spatial resolution of 8 and 80 mm respectively.
#| cache: true

@@

Before each flight, targets used as Ground Control Points (GCPs) were distributed over the study site and georeferenced with a Trimble © Geo XH 6000 differential GPS (dGPS). GCPs were used to correct georeferencing imprecision of orthomosaics with an horizontal and vertical accuracy of 10cm. dGPS was also used to georeferenced quadrats of 0.25 m² assessing the presence or absence of five key taxonomic classes of intertidal vegetation : Bacillariophyceae (Unicellular benthic diatoms forming biofilms at the sediment surface during low tide), Phaeophyceae (brown macroalgae), Magnoliopsida (dwarf seagrass), Chlorophyceae (green macroalgae) and Rhodophyceae (red macroalgae) (@fig-vegetation). Pictures of each quadrat were uploaded online to the Global Biodiversity Information Facility (GBIF) platform [@BedeGbif], a public open portal to store and share biodiversity data. Each photograph was also processed to estimate the percent cover of each type of vegetation using an image processing software (imageJ). For all quadrats, the hyperspectral reflectance signatures of each vegetation class was measured using an ASD FieldSpec HandHeld 2 spectroradiometer, which acquires reflectance between 325 and 1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures from quadrats serve for dual purposes: they validate the radiometric calibration of drone data and contribute to error reduction in photo interpretation. The workflow developed in this study is presented in @fig-workflow.

@@

```{r Table2}
#| echo: false
#| warning: false
#| tbl-cap: Vegetation Classes of the model and the number of pixels used to train and validate each class.
#| label: tbl-validationPX
#| cache: true

@@

A dataset containing photo-interpreted drone reflectance pixels was built to train a Neural Network model with 2 hidden layers. The training pixels were categorized into seven different classes, representing the various habitats encountered at the different study sites: Sediment, Water, Chlorophyceae, Magnoliopsida, Bacillariophyceae, Phaeophyceae and Rhodophyceae. Only low-altitude flights (@tbl-flights) were used for training because their 8 mm spatial resolution allowed to avoid spectral sub-pixel mixing and to accurately identify vegetation classes. More than `r my_comma(round(sum(df_training$Training_Pixels),-3))` pixels at 8 mm resolution from the 3 training flights were used to train the model (@tbl-validationPX). Twenty one variables were used by the model as predictors: the ten raw spectral bands of the Micasense RedEdge Dual MX multispectral camera (ranging from 444 nm to 840 nm), the same ten spectral bands standardized using a min/max transformation ([@eq-std] ; @Cao2017) and the Normalized difference vegetation index (NDVI, @eq-ndvi). Standardization of spectral bands is commonly used to eliminate the scaling differences between spectra and to limit the effect of biomass on the spectra shape [@Douay2022 ; @Davies2023].

@@

The model was applied to all flights at both 12 and 120 m of altitude. In situ information on georeferenced class type and percent cover collected during each flight was used to assess the model accuracy. These images were used to construct a validation dataset indicating the presence or absence of each class. Additionally to the quadrat-based validation dataset, polygons of each class were photo interpreted in order to increase the number of pixels of the validation dataset. A confusion matrix, along with precision metrics such as global accuracy, sensitivity, specificity, and Kappa coefficient, was generated for each site. All validation matrices were then aggregated to create a unique matrix. Altogether, a total of `r my_comma(round(count_validation_pixel,-3))` pixels was used to validate the model, thus providing a geographically robust validation dataset.

@@

The high-altitude flight over Gafanha covered a total area of approximately 1 km² (@fig-GafHigh). A channel contouring a small island was masked in the prediction map. Most of the intertidal area was classified as Magnoliopsida by the model, including seagrass patches with discolored leaves. Only a few pixels were classified as Chlorophyceae at this spatial scale. Furthermore, the area that was classified as Bacillariophyceae in the low-altitude flight remained mostly classified as such in the high-altitude flight, though some pixels were classified as Magnoliopsida. Patches of Rhodophyceae were correctly classified. In the northern part of the site and near the land edges, patches of the schorre angiosperm *Spartina sp.* were misclassified, either as Magnoliopsida or as Phaeophyceae.

@@

Among the high altitude flights, the one acquired over the inner lagoon of the Ria de Aveiro covered the largest area with approximately 1.5 km² (@fig-Boat). This site was dominated by seagrass and the red macroalgae *Gracilaria vermicullophyla*. The classification provided consistent results, with a patchy seagrass meadow mixed with red macroalgae on the eastern part of the site. As shown in the zoom (@fig-Boat), the edges of the meadow were colonized by green macroalgae

@@

```{r VIP Plot}
#| cache: false
#| echo: false
#| warning: false
#| fig-cap: Variable Importance of the Neural Network Classifier for each taxonomic class. The bigger the slice, the more important the variable for prediction of each class. The right plot shows the drone raw and standardized reflectance spectra of each class. Each slice represents the VIP of both raw and standardized reflectance combined.      
#| label: fig-VIP
#| fig-width: 10
#| fig-height: 7
#| out-width: "100%"

@@

```{r Upscaling Plot}
#| cache: false
#| echo: false
#| warning: false
#| fig-cap: Kernel density plot showing the proportion of pixel well classified based on the percent cover of the class in high altitude flight pixels of Gafanha, Portugal. Each subplot shows all the pixels of the same classes on the high altitude flight. Percent cover of classes was retrieved using the result of the classification of the low altitude flight of Gafanha, Portugal. The vertical dashed line shows the 0.85 probability of the model. Everything on the right of this line has a probability higher than 0.85 and everything on the left of this line has a lower probability.
#| label: fig-upscaling
#| fig-width: 10
#| fig-height: 7
#| out-width: "90%"

@@

The primary objective of this study was to develop a method for the accurate classification of macrophytes on intertidal mudflats, specifically focusing on distinguishing between Chlorophyceae (green macroalgae) and marine Magnoliopsida (seagrasses) using multispectral drone data. The ability to differentiate between various types of vegetation, as demonstrated in this study, plays a critical role in ecological monitoring and coastal management. By distinguishing between seagrasses and macroalgae, our approach facilitates targeted conservation strategies, enabling more effective preservation and restoration efforts in coastal ecosystems. The discrimination of seagrasses from green macroalgae presents significant challenges [@oiry2021using ; @bannari2022 ; @veettil2020opportunities], primarily due to the similarities on their spectral signatures in the visible region of the electromagnetic spectra due to shared pigment [@mckenzie2020global ; @fig-Pigm], but also by the frequent spatial mixing of these macrophytes. Besides chlorophyll-a that is common to all vegetation types, these two green macrophytes have in common other photosynthetic pigments such as chlorophyll-b, and accessory carotenoid pigments such as zeaxantin, lutein and neoxanthin (@fig-Pigm).

@@

This spectral overlap is further complicated by the intermingling of seagrasses and green macroalgae in the same spatial locations, posing a significant challenge for remote sensing aimed at accurately mapping and monitoring coastal ecosystems. Our study addresses these complexities by using high-resolution drone imagery with 10 spectral bands. @Davies2023 have shown that having at least eight spectral bands ranging between 500 nm to 850 nm including a green band at 530 nm and another one in the RedEdge region at 730 nm was crucial to accurately discriminate among green macrophytes.

@@

::: column
Meeting these two criteria, the Micasense RedEdge-MX DUAL camera used in this study, enabled the classifier to achieve `r round(validation_metrics_green$overall[[1]],2)*100`% of accuracy between these two classes (@fig-ValidationGreen). Even if the pigment composition of green macrophytes is similar, differences in the spectral shape can still be observed (@fig-vegetation). Several factors can explain these differences such has different concentration and/or pigment ratios [@bargain2013seasonal], different cellular organization and the disposition of the plant over the sediment surface [@beach1997vivo ; @kirk1994light ; @hedley2018influence].
:::
:::

The VIP analysis of the Neural Network model (@fig-VIP) identified that the band at 531 nm was the most important one for accurately identifying Chlorophyceae. In fact, at this wavelength, Chlorophyceae exhibited the highest reflectance (raw and standardized) among all other classes, highlighting the potential differences in accessory pigment ratios (i.e. carotenoid) between seagrasses and green macroalgae [@repolho2017seagrass].

Concerning Phaeophyceae, the thick cell walls of these macroalgae [@charrier2021growth] make it more reflective in the infrared part of the spectra [@Slaton2001] whereas the presence of Fucoxanthin and Zeaxanthin result in a low reflectance in the visible region (@fig-Pigm ; @fig-VIP). These two key features have been identified by the Neural Network as the two principal predictors to accurately identify brown algae (@fig-VIP). Similarly, the presence of phycoerythrin and phycocyanin in Rhodophyceae contributes to the lowest reflectance among all classes in the spectral range of 560 to 615 nm (@fig-VIP). Indeed the band at 560 nm has been identified as important for identifying this class likely due to phycoerythrin absorption at this wavelength (**REF**). Regarding Bacillariophyceae, the VIP analysis (@fig-VIP) indicated that 475 nm was the most important predictor for this class. This is not a usual diagnostic wavelength for Bacillariophyceae which is mainly identified by the presence of fucoxanthin and chlorophyll c (@fig-Pigm), but it correspond to the absorption of diadinoxanthin an accessory pigment characteristic of this class [@meleder2003spectrometric]. Furthermore, it is the vegetation with the lowest concentration of chlorophyll-a, pigment absorbing light both in the blue and the red. The transparency of Bacillariophyceae makes the reflectance of the sediment part of the overall reflectance of Bacillariophyceae, further explaining the high reflectance in the blue.

@@

While comparing the reflectance of both altitudes (12 m and 120 m), it was observed @fig-CompareRef that there is an underestimation of the infrared part of the spectra in the high-altitude dataset. Such disparity in infrared reflectance may stem from temporal differences between the flights, possibly resulting in a slightly drier intertidal area and consequently higher infrared reflectance. This disparity poses an issue for the methodology followed in the present study, relying solely on one flight height for training. To address this issue, we employed min/max standardized reflectance spectra as predictors for the model (@eq-std ; [@Cao2017]). This approach allowed us to eliminate the slight reflectance difference between the flights (@fig-CompareRef B) and to focus on the shape of the spectra in the visible part of the electromagnetic spectra, where different pigmentation are associated to taxonomic diagnostic features. This was a key feature in building a model that could reliably predict vegetation across geographical sites and seasons. It enabled consistent prediction of vegetation classes across variations in biomass and variability in light conditions [@fyfe2003spatial ; @COSTA2021107018 ; @piaser2023impact].
